---
title: "DOMS_with_flocker"
format:
  html:
    grid:
      body-width: 1100px
editor: visual
---

## Fit DOMs with flocker

Fit dynamic occupancy models with flocker package.

## Install / load packages

```{r}
#| output: false
#| warning: false
#install.packages("flocker", dependencies = TRUE)
library(flocker)
#install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
library(cmdstanr)
set_cmdstan_path(path = "C:/Users/schifferle1/Documents/cmdstan-2.34.1")
#check_cmdstan_toolchain() # requires RTools installation, see https://mc-stan.org/cmdstanr/articles/cmdstanr.html and https://mc-stan.org/docs/cmdstan-guide/installation.html
#install_cmdstan(dir = "C:/Users/schifferle1/Documents") # is that right?
library(brms)
library(dplyr)
```

## Load data

```{r}
# selected routes and focal years matched to environmental data:
# merged route, year, environment data:
load(file = file.path("data", "route_year_env_data.RData"))

# scale covariates:
route_sel_env_dt_scaled <- route_sel_env_dt_final %>% 
  mutate(across(bio2:pr_winter_3yrs, ~ (scale(.)) %>% as.vector()))

# route-year-species information (only surveyed)
load(file = file.path("data", "BBS_for_occ_spec_records.RData")) # output of 1_0_reformat_BBS_data.R

# selected species:
load(file = file.path("data", "final_species_selection.RData")) # output of 1_2_species_selection.R

```

Use House Finch as test species (fitting DOMs with JAGS went well for this).

```{r}
spec <- "House Finch"
```

### Assemble data for flocker

```{r}
#| output: false
nyears <- length(unique(route_sel_env_dt_final$Year)) # 25
nsurveys <- 5
nsites <- length(unique(route_sel_env_dt_final$RTENO)) # 476

# species presences:

presences_spec <- bbs_dt_occ %>% 
  select(c(English_Common_Name, RTENO, Year, paste0("Count", seq(10, 50, 10)))) %>% 
  filter(English_Common_Name == spec)

# match to routes-year-env:

occ_dt_spec <- route_sel_env_dt_scaled %>% 
  # add observations:
  collapse::join(presences_spec, on = c("RTENO", "Year"), how = "left") %>% 
  # if route was surveyed but species not observed, replace NA with 0:
  mutate(across(Count10:Count50, ~ 
                  case_when(Surveyed == 1 & is.na(.) ~ 0,
                            .default = .))) %>%
  # convert bird counts to presence / absence:
  mutate(across(Count10:Count50, ~ 
                  case_when(. > 1 ~ 1,
                            .default = .)))

# reformat obs. as array sites x surveys x years (same as for JAGS):

years <- seq(min(occ_dt_spec$Year), max(occ_dt_spec$Year))
y_array <- array(NA, dim = c(nsites, nsurveys, nyears))
for (t in 1:nyears){
  y_array[1:nsites, 1:nsurveys, t] <- as.matrix(occ_dt_spec[which(occ_dt_spec$Year == years[t]), c(paste0("Count", seq(10, 50, 10)))])
}

# reformat environmental covariates:

route_sel_env_dt_scaled
env_cov <- vector("list", length = nyears)
for (t in 1:nyears){
  env_cov[[t]] <- route_sel_env_dt_scaled[which(route_sel_env_dt_scaled$Year == years[t]), c("bio1", "bio2", "bio3", "pr_spring", "pr_summer","pr_autumn", "pr_winter", "bio15","bio1_3yrs", "bio2_3yrs", "bio3_3yrs", "pr_spring_3yrs", "pr_summer_3yrs", "pr_autumn_3yrs", "pr_winter_3yrs", "bio15_3yrs")]
  }

# covariate for detection probability:

det_cov <- vector("list", length = 1)
names(det_cov) <- "route_section"
det_cov$route_section <- array(NA, dim = c(nsites, nsurveys, nyears))
det_cov$route_section[ , , 1:nyears] <- matrix(rep(c("Sect1", "Sect2", "Sect3", "Sect4", "Sect5"), nsites), nsites, byrow = TRUE)

# make flocker data:

fd <- make_flocker_data_dynamic(
  obs = y_array,
  unit_covs = env_cov, 
  event_covs = det_cov, 
  quiet = FALSE
)

```

### Fit flocker model

#### Climate only with default priors:

```{r}
# multi_colex_cl <- flock(
#   f_occ = ~ bio1_3yrs + bio2_3yrs + bio3_3yrs + pr_spring_3yrs + pr_summer_3yrs + pr_autumn_3yrs + pr_winter_3yrs + bio15_3yrs +
#     I(bio1_3yrs^2) + I(bio2_3yrs^2) + I(bio3_3yrs^2) + I(pr_spring_3yrs^2) + I(pr_summer_3yrs^2) + I(pr_autumn_3yrs^2) + I(pr_winter_3yrs^2) + I(bio15_3yrs^2),
#   f_det = ~ 1,
#   f_col = ~ bio1 + bio2 + bio3 + pr_spring + pr_summer + pr_autumn + pr_winter + bio15 +
#     I(bio1^2) + I(bio2^2) + I(bio3^2) + I(pr_spring^2) + I(pr_summer^2) + I(pr_autumn^2) + I(pr_winter^2) + I(bio15^2) ,
#   f_ex = ~ bio1 + bio2 + bio3 + pr_spring + pr_summer + pr_autumn + pr_winter + bio15 +
#     I(bio1^2) + I(bio2^2) + I(bio3^2) + I(pr_spring^2) + I(pr_summer^2) + I(pr_autumn^2) + I(pr_winter^2) + I(bio15^2),
#   flocker_data = fd,
#   multiseason = "colex",
#   multi_init = "explicit",
#   backend = "cmdstanr",
#   cores = 3,
#   chains = 3
# ) # 54 min
# save(multi_colex_cl, file = file.path("data", paste0(gsub(" ", "_", spec), "_cl4_quadr_Stan.RData")))
load(file = file.path("data", paste0(gsub(" ", "_", spec), "_cl4_quadr_Stan.RData")))
multi_colex_cl

```

Is effective sample size large enough to get stable estimate of uncertainty? -\> Bulk_ESS = bulk effective sample size = how well is center of posterior distribution resolved. Tail_ESS = how well are extreme quantiles resolved (?).

```{r}
# extract estimates:
fixef(multi_colex_cl, summary = TRUE, robust = FALSE, probs = c(0.025, 0.975))
```

##### Get estimates for occupancy, extinction and colonisation probability:

For each site and year and survey (route section) we get mean and quantiles:

```{r}
# posterior predictive distribution for the modeled probabilities (occ., det., col., ext.):
# takes a while (20 min?)
# posterior_preds <- fitted_flocker( 
#   multi_colex_cl,
#   components = c("occ", "det", "col", "ex"),
#   new_data = NULL,
#   unit_level = FALSE,
#   summarise = TRUE,
#   CI = c(0.025, 0.975),
#   draw_ids = NULL,
#   response = TRUE, # response or logit scale
#   re_formula = NULL,
#   allow_new_levels = FALSE,
#   sample_new_levels = "uncertainty"
# )
#save(posterior_preds, file = file.path("data", paste0(gsub(" ", "_", spec), "_cl4_quadr_Stan_post_preds.RData")))
load(file = file.path("data", paste0(gsub(" ", "_", spec), "_cl4_quadr_Stan_post_preds.RData")))
str(posterior_preds)
posterior_preds$linpred_occ[1:10, , 1, ]

```

#### Modify priors:

Flocker gives us default priors for a model formula:

```{r}
get_flocker_prior(f_occ = ~ bio1_3yrs + bio2_3yrs + bio3_3yrs + pr_spring_3yrs + pr_summer_3yrs + pr_autumn_3yrs + pr_winter_3yrs + bio15_3yrs +
     I(bio1_3yrs^2) + I(bio2_3yrs^2) + I(bio3_3yrs^2) + I(pr_spring_3yrs^2) + I(pr_summer_3yrs^2) + I(pr_autumn_3yrs^2) + I(pr_winter_3yrs^2) + I(bio15_3yrs^2),
   f_det = ~ 1,
   f_col = ~ bio1 + bio2 + bio3 + pr_spring + pr_summer + pr_autumn + pr_winter + bio15 +
     I(bio1^2) + I(bio2^2) + I(bio3^2) + I(pr_spring^2) + I(pr_summer^2) + I(pr_autumn^2) + I(pr_winter^2) + I(bio15^2),
   f_ex = ~ bio1 + bio2 + bio3 + pr_spring + pr_summer + pr_autumn + pr_winter + bio15 +
     I(bio1^2) + I(bio2^2) + I(bio3^2) + I(pr_spring^2) + I(pr_summer^2) + I(pr_autumn^2) + I(pr_winter^2) + I(bio15^2),
                  flocker_data = fd,
                  multiseason = "colex",
                  multi_init = "explicit")
```

To change priors we use the brms implementation. Brms includes two shrinkage/ regularization priors, horseshoe and R2D2.

-   horseshoe:

    -   "Horseshoe prior is symmetric around zero with fat tails and an infinitely large spike at zero. This makes it ideal for sparse models that have many regression coefficients, although only a minority of them is non-zero."

    -   uses two levels of regularization, global and local, the global parameter τ pulls all the weights globally towards zero, while the thick half-Cauchy tails for the local scales λj allow some of the weights to escape the shrinkage

    -   brms arguments (regularized horseshoe):

        -   df (default 1): Degrees of freedom of student-t prior of the local shrinkage parameters

        -   scale_global (default 1): Scale of the student-t prior of the global shrinkage parameter; determines how much shrinkage is applied, Piironen and Vehtari (2017) recommend to specifying the ratio of the expected number of non-zero coefficients to the expected number of zero coefficients \`par_ratio\` rather than \`scale_global\` directly

        -   df_global (default 1): Degrees of freedom of student-t prior of the global shrinkage parameter

        -   scale_slab (default = 2): scale of student-t-slab (regularization, only affects non-zero coeff., expected value of the non-zero coefficient?)

        -   df_slab (default = 4): df of the student-t-slab (regularization, only affects non-zero coeff.)

        -   par_ratio: Ratio of the expected number of non-zero coefficients to the expected number of zero coefficients; specified, `scale_global` is ignored and internally computed as `par_ratio / sqrt(N)`, `N` is the total number of observations in the data

For now, I defined par_ratio instead of scale_global. I assumed that roughly 1/4 of the coefficients may be non-zero -\> par_ratio = (1/4)/(3/4) = 0.33. This results in a scale_global of 0.3/sqrt(11900) = 0.003 (default is 1). Not sure about this.

"Specifying priors for parameters in formulas other than detection can be done with reference to the `dpar` column."

```{r}
# multi_colex_all_cl_prior <- flock(
#   f_occ = ~ bio1_3yrs + bio2_3yrs + bio3_3yrs + pr_spring_3yrs + pr_summer_3yrs + pr_autumn_3yrs + pr_winter_3yrs + bio15_3yrs +
#     I(bio1_3yrs^2) + I(bio2_3yrs^2) + I(bio3_3yrs^2) + I(pr_spring_3yrs^2) + I(pr_summer_3yrs^2) + I(pr_autumn_3yrs^2) + I(pr_winter_3yrs^2) + I(bio15_3yrs^2),
#   f_det = ~ 1,
#   f_col = ~ bio1 + bio2 + bio3 + pr_spring + pr_summer + pr_autumn + pr_winter + bio15 +
#     I(bio1^2) + I(bio2^2) + I(bio3^2) + I(pr_spring^2) + I(pr_summer^2) + I(pr_autumn^2) + I(pr_winter^2) + I(bio15^2) ,
#   f_ex = ~ bio1 + bio2 + bio3 + pr_spring + pr_summer + pr_autumn + pr_winter + bio15 +
#     I(bio1^2) + I(bio2^2) + I(bio3^2) + I(pr_spring^2) + I(pr_summer^2) + I(pr_autumn^2) + I(pr_winter^2) + I(bio15^2),
#   flocker_data = fd2_det,
#   prior = c(brms::set_prior("logistic(0,1)", class = "Intercept") + # flat on probability scale (https://cran.r-project.org/web/packages/flocker/vignettes/flocker_tutorial.html)
#             brms::set_prior("logistic(0,1)", class = "Intercept", dpar = "occ"),
#             brms::set_prior("logistic(0,1)", class = "Intercept", dpar = "colo"),
#             brms::set_prior("logistic(0,1)", class = "Intercept", dpar = "ex"),
#             brms::set_prior("horseshoe(df = 1, par_ratio = 0.3)", dpar = "occ", class = "b"),
#             brms::set_prior("horseshoe(df = 1, par_ratio = 0.3)", dpar = "colo", class = "b"),
#             brms::set_prior("horseshoe(df = 1, par_ratio = 0.3)", dpar = "ex", class = "b")),
#   multiseason = "colex",
#   multi_init = "explicit",
#   backend = "cmdstanr",
#   cores = 3,
#   chains = 3
# )# 11 of 3000 transitions divergent, 12000s = 3.3 hrs
# save(multi_colex_all_cl_prior, file = file.path("data", paste0(gsub(" ", "_", spec), "_cl4_quadr_Stan_prior.RData")))
load(file = file.path("data", paste0(gsub(" ", "_", spec), "_cl4_quadr_Stan_prior.RData")))
multi_colex_all_cl_prior

```

##### Compare to same model in JAGS using lasso regularization:

```{r}
library(jagsUI)
load(file.path("results", "House_Finch_cl_alt4_quadr_lasso_GoF_inits.RData"))

jags_estimates <- out$summary[which(grepl("(alpha)|(beta)", rownames(out$summary))), c("mean", "2.5%", "97.5%")]

stan_estimates <- fixef(multi_colex_all_cl_prior, summary = TRUE, robust = FALSE, probs = c(0.025, 0.975))

jags_reformatted <- rbind(c(NA,NA, NA),
                          jags_estimates["alpha_psi", c("mean", "2.5%", "97.5%")],
                          jags_estimates["alpha_gamma", c("mean", "2.5%", "97.5%")],
                          jags_estimates["alpha_eps", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_psi[1]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_psi[2]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_psi[3]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_psi[4]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_psi[5]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_psi[6]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_psi[7]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_psi[8]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_psi[9]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_psi[10]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_psi[11]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_psi[12]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_psi[13]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_psi[14]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_psi[15]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_psi[16]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_gamma[1]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_gamma[2]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_gamma[3]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_gamma[4]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_gamma[5]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_gamma[6]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_gamma[7]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_gamma[8]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_gamma[9]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_gamma[10]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_gamma[11]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_gamma[12]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_gamma[13]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_gamma[14]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_gamma[15]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_gamma[16]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_eps[1]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_eps[2]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_eps[3]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_eps[4]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_eps[5]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_eps[6]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_eps[7]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_eps[8]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_eps[9]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_eps[10]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_eps[11]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_eps[12]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_eps[13]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_eps[14]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_eps[15]", c("mean", "2.5%", "97.5%")],
                          jags_estimates["beta_eps[16]", c("mean", "2.5%", "97.5%")])

comparison <- cbind(as.data.frame(stan_estimates)[,c(1,3,4)], jags_reformatted)
colnames(comparison) <- c("Stan_mean", "Stan_2.5%", "Stan_97.5%",
                          "JAGS_mean", "JAGS_2.5%", "JAGS_97.5%")
comparison <- comparison[,c(1,4,2,5,3,6)]
comparison
```

Quite similar.

#### Post-processing:

Posterior distribution for the latent occupancy state / occupancy probabilities across sites (option to do this for new data):

```{r}
#lat_occ <- get_Z(multi_colex_all_cl_prior) # takes a while
#save(lat_occ, file = file.path("data", paste0(gsub(" ", "_", spec), "_cl_lat_occ_hs.RData")))
load(file = file.path("data", paste0(gsub(" ", "_", spec), "_cl_lat_occ_hs.RData")))
str(lat_occ) # for each site and year we have 3000 draws
lat_occ[241,3,][1:10]

```

Posterior predictions at the observed points (e.g. for use in posterior predictive checking) or for new data:

```{r}
#post_preds <- predict_flocker(multi_colex_all_cl_prior) 

```

Model comparison:

The "leave-one-out" holdouts consist of entire closure-units (single-season models), series (multi-season models).

```{r}
#loo_mod <- loo_flocker(multi_colex_all_cl_prior, thin = 2)
# Warning: Some Pareto k diagnostic values are too high. See help('pareto-k-diagnostic') for details
#save(loo_mod, file = file.path("data", paste0(gsub(" ", "_", spec), "_cl_loo_hs.RData")))
load(file = file.path("data", paste0(gsub(" ", "_", spec), "_cl_loo_hs.RData")))
loo_mod
#loo::pareto_k_table(loo_mod)
loo::pareto_k_ids(loo_mod) # 419, 422 = sites above threshold, highly influential?

# values <- loo::pareto_k_values(loo_mod)
# infl_values <- loo::pareto_k_influence_values(loo_mod) # influence of observations on model posterior distribution
# identical(values, infl_values) # TRUE

#loo_mod2 <- loo_flocker(multi_colex_all_cl_prior, thin = NULL)
#save(loo_mod2, file = file.path("data", paste0(gsub(" ", "_", spec), "_cl_loo2_hs.RData")))
load(file = file.path("data", paste0(gsub(" ", "_", spec), "_cl_loo2_hs.RData")))
loo_mod2
```

What does this mean?

https://mc-stan.org/loo/reference/loo-glossary.html

-   elpd = theoretically expected log pointwise predictive density for a new dataset, can be estimated using cross-validation

-   elpd_loo = The Bayesian LOO estimate of out-of-sample predictive fit

    -   predictive density given the data without the iths data point/series, log transformed and summed over all data points/series

-   p_loo = difference between elpd_loo and non-cross-validated log posterior predictive density, how much more difficult is it to predict future data compared to observed data

    -   can be interpreted as effective number of parameters, if everything is fine: p_loo \< N and p_loo \< p (=total number of parameters)

        -   warning: Some Pareto k diagnostic values are too high. See help('pareto-k-diagnostic') for details:

-   looic = -2\*elpd_loo

-   Pareto k diagnostic estimates how far an individual leave-one-out distribution is from the full distribution

    -   diagnostic threshold for Pareto k depends on sample size S, `loo` package uses the nominal sample size S when computing the sample size specific threshold

    -   Pareto k is also useful as a measure of influence of an observation. Highly influential observations have high k values.

```{r}
#loo_comp <- loo_compare_flocker(model_list = list(multi_colex_cl, multi_colex_all_cl_prior), thin = 2)
#save(loo_comp, file = file.path("data", paste0(gsub(" ", "_", spec), "_cl_loo_comp_def_hs.RData"))) #Warning: Some Pareto k diagnostic values are too high. See help('pareto-k-diagnostic') for details
load(file = file.path("data", paste0(gsub(" ", "_", spec), "_cl_loo_comp_def_hs.RData")))
loo_comp

```

This tells me that model2 = model with regularizing priors is better (is this a large difference? what about the warning?)

Divergent transitions:

-   If you get divergence warnings, geometry of posterior density does not allow finding a single suitable step size. Posterior density cannot be sampled well. Clear signal that something went wrong. Increasing target average acceptance probability (delta) may help.

Notes:

-   there is nothing like autojags that increases iterations until chains have mixed

-   you can't explicitly state which parameters to monitor. It looks like only estimates for coefficients are monitored (not the intermediate responses)
